paper1_figures

### 2023-12-12 scoping_review_short.Rmd
12.12: integration on Vincent's and Lilla's basic feedbacks

---
title             : "Improving the understanding of cooperation in commons through social dilemmas' literature: A scoping review"
shorttitle        : "Improving the understanding of cooperation in commons"

@@

abstract: |
  Commons represent an alternative to the classical capitalist model, therefore they could be the needed change toward a more sustainable society and economy. However, in order for a commons to be successful, its members must cooperate, i.e. incur a personal sacrifice for the collective good. Despite the importance of cooperation, there is a lack of studies that investigated this topic in real-life commons. On the other hand, cooperation has been extensively studied in social dilemma experiments, where personal and collective interests are at odds. Since the two contexts share some structural characteristics, the results of social dilemma experiments could be used to gain insights on cooperation in commons.
  The goal of the current scoping review is to identify the factors that impact cooperation in social dilemma situations, in order to better understand what could shape cooperation in commons. A key-word search and a subsequent selection lead to the inclusion of 89 papers, who are used to identify 9 factors that can influence cooperation both at the individual and group level: gender, social status, group identification, values and personality traits, trust, incentives, communication, social norms and anonymity. These factors are discussed within the existing literature on commons and cooperation. More specifically, the influences that these elements can have on each-other are highlighted, showing that the totality of factors should be taken into account when investigating cooperation in commons. Finally, the practical implication of this work for commons' organization, as well as the review's limitations, are pointed out.

@@

During last years, the interest in alternative ways of social organization increased drastically because of the social and ecological crises we are facing [@Euler2018]. Commons are one of these alternatives, since they are a way of organizing resources not involving the classic private and public model [@Ostrom1990], and, therefore, they represent an option to the capitalistic system [@Euler2019]. For this reason, commons could be the needed change to attain a more sustainable and fair social organization [@Seravalli2018].

Commons have been defined in different ways, starting by Elinor Ostrom who, in her seminal work, refers to them as a resource system shared between a self-managed community of users [@Ostrom1990]. This community has a set of social practices, values and norms. Therefore, it is important to consider the commons not only as a resource, but as the integration of the resource, the community and its social protocols [@Bollier2014]. For this reason, beyond the material aspect, the notion of social practices (i.e. commoning) is essential [@Euler2018]. "Commoning" refers to the process through which people (i.e. the commoners) find solutions for their needs by self-organizing independently from the state or the market, and by creating autonomously managed institutions to govern the shared resource [@Euler2018; @Schmelzer2022]. The community is therefore engaged in defining rules to successfully manage the commons [@Meroni2022]. In conclusion, commons can be understood as "living social systems" who allow commoners to self-organize to address shared problems [@Bollier2019].

These definitions highlight the importance of the community as a part of the commons, and that is why it is essential to have a closer look to the people who form it. More specifically, in order for a commons[^1] to be successful, its members have to cooperate, i.e to incur a personal sacrifice for the greater good [@Kollock1998]. Although its importance, cooperation can be difficult to achieve due to two characteristics of the commons' social dynamic. First, people can believe that their individual contribution will not have a big impact, and this feeling of personal inefficacy discourages cooperation [@VanVugt1998]. Secondly, in a commons, every member faces a conflict between individual, short-term interests and collective, long-term ones [@Hardin1968]. Act for the common good requires a personal sacrifice, and therefore the temptation to pursue the individual interest is constantly present. For these reasons, it is essential to understand how cooperation in commons can be improved, and which factors impact it.

@@

There are therefore several studies that focused on commons and on how they can be successful, but social psychological and quantitative perspectives are missing from this literature. In fact, to the authors' knowledge, there are few or no studies that investigated the causal effect of individual and group factors on people's willingness to cooperate in a real-life commons.

At the same time, this approach is very popular in the laboratory study of social dilemmas, which allowed to identify several factors that shape cooperation [e.g. @VanLange2013]. In this experiments, people face a conflict between personal and collective interest: each individual obtains a higher personal payoff with a non-cooperative, self-interested behaviour than with a social, cooperative one, independently from what others do. At the same time, everybody's payoff is higher if all cooperate than if all defect [@Dawes1980]. Since many of the current global problems have a social dilemma structure, this topic has been extensively studied in different fields such as psychology, economy, biology, or mathematics [@VanLange2013]. Thanks to their easy implementation in a laboratory, social dilemmas are very useful to gain insights into how cooperation works, because they provide a standard experimental paradigm to test under which conditions people cooperate [@Balliet2011].

@@

Commons and social dilemmas are deeply linked. First, they share the same basic conflict between personal and collective interests [@Hardin1968; @Dawes1980]. For this reason, commons can be considered as a particular case of social dilemma. In fact, @Kopelman2002 used the term of "commons dilemma" to define situations where a non cooperative behaviour can seriously endanger the future of the shared resource. Moreover, commons have been classified by @Cumming2018 as "take-some dilemmas", a specific kind of social dilemma where people are tempted to increase their personal benefit by taking more than their fair share from a shared resource. Moreover, cooperation and defection in social dilemma laboratory experiments can predict people's behaviour in a real-life commons [e.g. @Reindl2019].

The aim of the current paper is to synthesize the existing researches about cooperation in social dilemmas, to provide a set of factors that could impact cooperative behaviours in commons. This work will therefore lay the foundations to implement experimental and quantitative studies in real-life commons, which, in turn, will allow to gain insights about cooperation in this context. Other studies already used a social dilemma framework to study real-life commons [e.g. @Hartl2022]. While these studies usually focused on just one particular case, the current review is seeking to extend this approach by using social dilemma literature to gain a global understanding of cooperation in commons.

This scoping review has two main goals. The first is to identify individual factors that can shape cooperation in commons. In fact, taking into account the specific community and commoners is essential to fully understand a commons' functioning [@Dove2019]. Moreover, since it has been widely demonstrated that commoners can successfully self-organize to govern a commons [@Ostrom1990; @Larsson2016], the second goal is to understand which strategies groups can implement to encourage their members to cooperate.

@@

To conduct this scoping review, different scientific publications were analyzed in response to the research questions. To gather the papers, we performed a Boolean search methodology in Scopus, Web of Science and ScienceDirect. The main key phrases were "social dilemma", "public good", "common resource", and "commons", and each of them was paired with each one of the following: "cooperation", "conflicts", "reciprocity", and "trust". The research was made during the month of August, 2023. According to @Munn2022, the type of source was not a selection criterion, therefore journal articles, conference papers and book chapters could be included, to have an idea as complete as possible about the existing literature on the topic. Moreover, there were not any chronological or geographical limitations, once again to provide a broader vision. There were however some selection criteria concerning the papers' methodology. First, only papers written in English were selected. Moreover, human participants were needed, so studies of social dilemmas through an agent-based model were excluded. Secondly, among experimental or correlative studies, only cases with adult participants were selected.

Initially, 5210 papers emerged from the key phrases search. Deleting the duplicates reduced them to 4586. Based on titles and abstracts, 4333 articles were eliminated, leaving 253 texts to examine. After a full-text evaluation of the remaining papers, 89 were selected for the present review. Figure 1 resumes this selection process.

@@

The 89 included papers were published between 1984 and 2023, with a median of 2015. Almost all of the selected texts were journal articles, with the only exception being a book chapter. Most of them were experimental studies (*N*=60), followed by literature review and meta-analysis (*N*=14) and correlative studies[^3] (*N*=14). The remaining paper was a qualitative study.

@@

The papers' topic were in four categories: first, "lab-based social dilemma" (*N*=75) refers to highly controlled laboratory experiments where participants had to take cooperative decision. The second category is "real-life social dilemma" (*N*=7), and investigated cooperation in different real-life situations with a social dilemma structure such as recycling behaviours [@Rompf2017] or use of public transportation [@VanLange1998]. Thirdly, papers in the category "lab-based commons" (*N*=3) used a laboratory social dilemma paradigm, but with clear references to a concrete common. For instance, in @Mosler1993, participants had to imagine to be fishermen who shared a common basin. Finally, "real-life commons" (*N*=4) refers to papers who studied cooperation with participants that were members of a commons in their everyday life.

@@

Most of the papers concerned psychology and economy. In fact, more than a half was published in psychology journals, such as *Group Processes & Integroup Relations* or *Psychological Bulletin* (*N*=34), in economic journals like *Experimental Economics* (*N*=14), or in interdisciplinary ones such as *Journal of Economic Psychology* (*N*=8).

@@

Finally, most of the studies were conducted in Europe (*N*=40) (especially in Germany (*N*=12) and in the Netherlands (*N*=8)), followed by North America (*N*=20, of which *N*=19 in the USA) and Asia (*N*=7, of which *N*=6 in China). Moreover, six studies were set in more than one country, and fourteen were meta-analysis or literature reviews. The two remaining studies took place in Australia and in Colombia, respectively.

@@

Five individual factors that can impact cooperation emerged from the literature review: gender, social status, group identification, values and personality traits, and trust. In the following sections, their principal effects on cooperation are described. Moreover, an overview of the moderators of each effect is provided. More details about moderators are included in the Appendix A.

@@

The effect of gender on cooperation is controversial, and studies that investigated it provided contradictory results. Sometimes men cooperated more [@Dorrough2019], sometimes women did [@Peshkovskaya2017], but two meta-analysis found a non significant effect size [@Balliet2011; @Spadaro2023]. This indicates that gender differences in cooperation are very flexible and depend on the context [@Spadaro2023]. More specifically, @Sell2009 explained this contradictory results "by structural differences [of the experimental settings] and identities triggered by those differences" (p. 322).

@@

Lastly, gender is also linked to social status, with women representing the low-status group [@Sell2009]. Therefore, gender differences in cooperation could also be partly explained by social status: to have a better understanding of this hypothesis, it is important to gain insights on how social status impacts cooperation.

@@

The results of studies investigating the impact of social status on cooperation do not allow to draw a clear conclusion about this effect. For instance, some findings suggest that low-status participants cooperate more than high-status ones by contributing a higher proportion of their endowments [@Kingsley2016]. However, @VanLange2013 identified some studies where wealthier people cooperated more.

Social status can also create inter-class conflicts, because people tend to cooperate more with the members of the same social class than with other classes' members [@Aksoy2019]. This is especially true for low-status groups, where solidarity can arise either in the form of discrimination toward high-status members [@Camera2020], or of less punishments for other low-status [@Chen2022].

Regarding the overall group composition, results are not consensual. On one hand, @Aksoy2019 found that groups with an inequality of status between members displayed lower levels of cooperation than homogeneous groups. On the other hand, economic and socio-cultural heterogeneity, although being negatively correlated with in-group trust, does not have a direct effect on a common-pool resource well-being [@VanKlingeren2021].

Lastly, similarly to what @Sell2009 affirmed about the effects of gender and social status, @Aksoy2019 demonstrated that the lower level of cooperation between members of different social classes is mainly due to actors' identification with their classes. This shows that social identity and social status are deeply connected, and that it is difficult to separate their effects on cooperation.

@@

In general, people cooperate more with other in-group members than with out-group [@Dorrough2015; @Aksoy2019], but the individual differences in the strength of identification can lead to inequalities on cooperation. High-identified individuals can even compensate, by cooperating more, for other in-group members' defection [@VanVugt2009; @Arora2016].

Intragroup cooperation can be enhanced via a competition between different groups, because in such a context individuals' and groups' interests are aligned [@Puurtinen2009]. Moreover, a study showed that intergroup competition increased cooperation, while individual, dispositional competitiveness had the opposite effect [@Nockur2020]. This demonstrated that the increase of cooperation via competition is due to group membership factors, and not to competition in itself.

Despite the beneficial effects of inducing an intergroup competition, this could be harmful in real-life situations, like resources shared between several groups. In fact, there is a greater risk of over-exploitation if an intergroup competition is present [@VanVugt2009]. To maintain the positive effect of group identification while eliminating the risks associated to the competition, it is useful to prime a super-ordinate identity [@VanVugt2009]. This can be done by making salient the common characteristics of the different groups rather than their differences [@Kramer1984] or by inducing the feeling of a common fate [@Zhang2019].

@@

Trust can be defined as the "expectations of others' benevolent motives in situations that involve a conflict between self and collective interests" [@Balliet2013, p.1]. There are two category of trust: first, dispositional trust is an individual general trait that can be assessed with personality scale. Secondly, state trust is linked to the context and represent the expectations about a partner's behaviour in a specific situation [@Balliet2013]. The two are positively correlated [@Lubke2021], and both have a positive general effect on cooperation [@VanLange2013; @Balliet2013].

Several studied confirmed this relationship. For instance, high levels of trust are associated with more contributions to a public good [@Kocher2015], individuals' cooperation can be predicted by their beliefs about others' contributions [@Smith2013], people who are pessimistic about others' contribution cooperate significantly less than optimistic [@Oyediran2018], and individuals who cooperate more display higher levels of trust in following interactions [@Chaudhuri2002]. The effect of trust on cooperation has also been demonstrated in real-life studies [e.g. @Sturm2019; @Franzen2019]. Finally, trust matters more in situations involving a large, rather than small, conflict between personal and collective interests [@Balliet2013], and when people are uncertain about others' intentions [@VanLange2013].

People can also display different levels of trust in the "reliability, effectiveness and legitimacy of public institutions" [@Rompf2017, p. 2]. Institutional trust is associated with the probability of recycling, because knowing that sanctions can be efficiently applied motivate people to comply to norms.

@@

On the organizational level, four factors have emerged from the literature review: incentives, communication, social norms and anonymity. As for the individual factors, the main effects of each factor are detailed in the next sections, as well as an overview of possible moderators. More details can be found in Appendix A.

@@

Cooperation can be modified by introducing incentive systems, that can be either punishments or rewards. Both have a positive effect on cooperation [@Balliet2011a; ]. However, there are contradictory results about which system is more efficient [see @Chen2022; @Milinski2012].

Several factors moderate this main effect. A meta-analysis found that both rewards and punishments are more effective if their implementation is costly rather than free, and during iterated, rather than single, interactions [@Balliet2011a]. Moreover, punishments work better in groups where all members have the same income [@Kingsley2016], because in heterogeneous groups a form of solidarity between low-status members arises and encourages free-riding [@Chen2022]. Punishments are also more effective if participants are friends [@Balliet2011a] and when only cooperators have the power to punish defectors [@Xiao2016]. Finally, the link between behaviour and outcome must be clear [@Xiao2016], and the punishment must have a significant consequence [@Egas2008].

In general, there are more studies about punishments than about rewards [@VanDijk2015]. When a punishment system is present, participants' perceived fairness is higher [@Nockur2021]. However, punishments must be applied with carefulness. Ideally, groups should not punish in early stages of the interaction, and have a norm of when and how punish [@Rockenbach2019]. Moreover, punishment could entails a second-order social dilemma, since cooperators are better-off by non punishing, if all others do [@Milinski2012]. Also, if there are no rules on who can punish, non-cooperators can punish cooperators as a form of retaliation [@Xiao2016; @VanDijk2015]. A democratic punishment system, where a fine is given only if a majority agree, could solve these problems [@Pfattheicher2018].

Even if they enhance cooperation, incentives have negative sides: they reduce participants' intrinsic motivation [@VanDijk2015; @Schlosser2018] by framing the situation as an economic, rather than ethical, dilemma [@VanVugt2009]. They also undermine participants' trust in other members' internal motivation to cooperate. Therefore, if they are removed, cooperation drops [@Irwin2014; @Mulder2006].

Finally, incentives are often studied in economic formats, but they can also be social. In fact, expecting honour or shame (after cooperating or defecting, respectively), or expressing disapproval toward defectors can enhance cooperation [@Jacquet2011; @Nelissen2013]. Even an ostracism system where participants can vote to exclude other members is very effective [@Maier-Rigaud2010]. In conclusion, social sanctions can be even more efficient than economic ones, because cooperation do not drop after their removal [@Nelissen2013].

@@

Communication between members has a global positive effect on cooperation [e.g. @Balliet2010], because it allows to shape a normative content [@Janssen2014; @Bicchieri2002] and to reach a consensus about the best strategy [@Hopthrow2005]. In fact, when communication is used to foster a common social identity, promote cooperative social norms and ensure that every member publicly commits, communication is higher [@Adams2022].

Concerning the moderation effects, oral communication is more powerful than written messages, and communication has a stronger impact in large, rather than small, groups [@Balliet2010]. Moreover, if communicating requires a personal cost, its effect is weaker than if communication is free [@Janssen2014]. The content of the message is also essential: to be effective, communication must allow participants to become aware of the problem, identify different strategies and agree on the one to use. However, casual chit-chat also plays a very important role in increasing cooperation [@Koessler2021]. Finally, communication before or during the cooperative interaction have both a positive impact on cooperation [@Balliet2010], while knowing that it will be possible to communicate after the interaction does not enhance cooperation [@Kumakawa2013], and can even have a negative effect [@Torsvik2011].

A particular form of communication is gossip, i.e. the spread of reputational information about another person [@Feinberg2014]. Gossip increases cooperation by providing information about the past behaviour of potential partners, which helps to select the most cooperative members in a following interaction. Moreover, knowing that gossip is possible motivates people to cooperate more to avoid being the target of such gossip [@Feinberg2014]. Gossip can also have the pro-social function of protecting others from a potential exploitation [@Feinberg2012]. Finally, providing correct information while gossiping helps to build a positive reputation [@Giardini2021] and reduces negative affects on the gossip actor [@Feinberg2012].

@@

Social norms can provide a guide on how to behave, especially when there is uncertainty about others' behaviour like in cooperative situations [@VonBorgstede2018]. Both injunctive (what others approve) and descriptive (what others do) norms are positively related to cooperative behaviours [@Thogersen2008]. Moreover, social norms help to maintain cooperation, because individuals who do not respect them are susceptible of social sanction [@Rege2004].

>>>>

---
title             : "Understanding cooperation in commons through social dilemmas' literature: a scoping review"
shorttitle        : "Understanding cooperation in commons"

@@

abstract: |
  Commons represent an alternative to the classical capitalist model, therefore they could be the needed change toward a more sustainable society and economy. However, for a commons to be successful, its members must cooperate, i.e. incur a personal sacrifice for the collective good. Despite the importance of cooperation, there is a lack of studies that investigated this topic in the scientific field of the commons. On the other hand, cooperation has been extensively studied in social dilemma experiments, where personal and collective interests are at odds. Since social dilemma experiments and commons share some structural characteristics, we could gain insights into cooperation in commons and how to implement it thanks to the social dilemmas literature.
  The goal of the current scoping review is to identify the factors that impact cooperation in social dilemma situations, in order to better understand what could shape cooperation in commons. A key-word search and a subsequent selection led to the inclusion of 89 papers, which were used to identify 9 factors that can influence cooperation both at the individual and group level: gender, social status, group identification, values and personality traits, trust, incentives, communication, social norms and anonymity. These factors are discussed within the existing literature on commons and cooperation. More specifically, the influences that these elements can have on each-other are highlighted, showing that all factors should be taken into account when investigating cooperation in commons. Finally, the practical implication of this work for commons' organization, as well as the review's limitations, are pointed out.

@@

In recent years, the interest in alternative ways of social organization increased drastically due to the social and ecological crises we are facing [@Euler2018]. Commons are one of these alternatives, since they are a way of organizing resources beyond the classic private and public model [@Ostrom1990], and, therefore, they represent an option to the capitalistic system [@Euler2019]. For this reason, commons could be an integral part of the path to attain a more sustainable and fair social organization [@Schmelzer2022].

Commons have been defined in different ways, starting by Elinor Ostrom who, in her seminal work, refers to them as a resource system shared between a self-managed community of users [@Ostrom1990]. This community shares a set of social practices, values and norms. Therefore, it is important to consider the commons not only as a resource, but as the integration of the resource, the community and its social protocols [@Bollier2014]. For this reason, beyond the material aspect, the notion of social practices (i.e. commoning) is essential [@Euler2018]. "Commoning" refers to the process through which people (i.e. the commoners) find solutions for their needs by self-organizing independently from the state or the market, and by creating autonomously managed institutions to govern the shared resource [@Euler2018; @Schmelzer2022]. The community is therefore engaged in defining rules to successfully manage the commons [@Meroni2022]. In conclusion, commons can be understood as "living social systems" that allow commoners to self-organize to address shared problems [@Bollier2019].

These definitions highlight the importance of the community as a part of the commons, and that is why it is essential to have a closer look at the people who form it. More specifically, in order for a commons[^1] to be successful, its members have to cooperate, i.e to incur a personal sacrifice for the greater good [@Kollock1998]. Although it is essential, cooperation can be difficult to achieve due to two characteristics of the commons' social dynamic. First, people can believe that their individual contribution will not have a big impact, and this feeling of personal inefficacy discourages cooperation [@VanVugt1998]. Secondly, in a commons, every member faces a conflict between individual, short-term interests and collective, long-term ones [@Hardin1968]. Act for the common good requires a personal sacrifice, and therefore the temptation to pursue the individual interest is constantly present. For these reasons, we need to understand the factors that affect cooperation, and how we can improve it.

@@

There are therefore several studies that focused on commons and on how they can be successful, but social psychological and quantitative perspectives are missing from this literature. In fact, to the authors' knowledge, there are few studies that investigated the causal effect of individual and group factors on people's willingness to cooperate in a real-life commons [e.g. @Noonan2016].

At the same time, an experimental approach is very popular in the laboratory study of social dilemmas, which identified several factors that shape cooperation [e.g. @VanLange2013]. In these experiments, people face a conflict between personal and collective interest: each individual obtains a higher personal payoff with a non-cooperative, self-interested behaviour than with a social, cooperative one, independently from what others do. At the same time, everybody's payoff is higher if all cooperate than if all defect [@Dawes1980]. Since many of the current global problems have a social dilemma structure, this topic has been extensively studied in different fields such as psychology, economy, biology, or mathematics [@VanLange2013]. Social dilemmas are very useful to gain insights into how cooperation works, because they can easily be implemented in a laboratory setting, and they provide a standard experimental paradigm to test under which conditions people cooperate [@Balliet2011].

@@

Commons and social dilemmas are deeply linked. First, they share the same basic conflict between personal and collective interests [@Hardin1968; @Dawes1980]. For this reason, commons can be considered as a particular case of social dilemma. In fact, @Kopelman2002 used the term of "commons dilemma" to define situations where a non-cooperative behaviour can seriously endanger the future of the shared resource. Moreover, commons have been classified by @Cumming2018 as "take-some dilemmas", a specific kind of social dilemma where people are tempted to increase their personal benefit by taking more than their fair share from a shared resource. Moreover, cooperation and defection in social dilemma laboratory experiments can predict people's behaviour in a real-life commons [e.g. @Reindl2019].

The current paper synthesizes existing research about cooperation in social dilemmas, to provide a set of factors that impact cooperative behaviours in commons. This work will therefore lay the foundations to implement experimental and quantitative studies in real-life commons, which, in turn, will allow to gain insights about cooperation beyond artificial laboratory settings. Other studies already used a social dilemma framework to study real-life commons [e.g. @Hartl2022]. While these studies usually focused on just one particular case, the current review is seeking to extend this approach by using social dilemma literature to gain a global understanding of cooperation in commons.

This scoping review has two main goals. The first is to identify factors on the individual level that can shape cooperation in commons. Moreover, since it has been widely demonstrated that commoners can successfully self-organize to govern a commons [@Ostrom1990; @Larsson2016], the second goal is to understand which strategies groups can implement to encourage their members to cooperate.

@@

To conduct this scoping review, we analyzed different scientific publications in response to the research questions. To gather the papers, we performed a Boolean search in Scopus, Web of Science and ScienceDirect. The key phrases were "social dilemma", "public good", "common resource", and "commons", and each of them was paired with each one of the following: "cooperation", "conflicts", "reciprocity", and "trust". We carried out the research during the month of August, 2023. According to @Munn2022, the type of source was not a selection criterion, therefore journal articles, conference papers and book chapters could be included, to have an overview as complete as possible about the existing literature on the topic. Moreover, we did not use any chronological or geographical limitations, once again to provide a broader vision. However, we implemented some selection criteria concerning the papers' methodology. First, we only selected papers written in English. Moreover, we focused on studies with human participants (or theoretical articles), so we excluded researches using an agent-based model. Secondly, among experimental or correlative studies, we only selected cases with adult participants.

Initially, 5210 papers emerged from the key phrases search. Deleting the duplicates reduced them to 4586. Based on titles and abstracts, we eliminated 4333 articles, leaving 253 texts to examine. After a full-text evaluation of the remaining papers, we selected 89 for the present review. Figure 1 resumes this selection process.

![Figure 1. Overview of the selection process.](images/figure1-01.png)

@@

The 89 included papers were published between 1984 and 2023, with a median of 2015. Almost all the selected texts were journal articles, with the only exception being a book chapter. Most of them were experimental studies (*N*=60), followed by literature review and meta-analysis (*N*=14) and correlative studies[^3] (*N*=14). The remaining paper was a qualitative study.

@@

We identified two dimensions in the reviewed papers: setting (lab-based vs real-life) and context (social dilemma vs commons). There were papers in all four combinations of these dimentions, with the most represented category being "lab-based social dilemma" (*N*=75), namely highly controlled laboratory experiments where participants had to take cooperative decision. This was followed by "real-life social dilemma" (*N*=7), which investigated cooperation in different real-life situations with a social dilemma structure such as recycling behaviours [@Rompf2017] or use of public transportation [@VanLange1998]. Finally, "real-life commons" refers to studies of cooperation where participants were members of a commons in their everyday life, while "lab-based commons" refers to studies with a laboratory social dilemma paradigm, but with clear references to a concrete commons [e.g. @Mosler1993].

@@

More than a half of papers was published in psychology journals, such as *Group Processes & Integroup Relations* or *Psychological Bulletin* (*N*=34), in economic journals like *Experimental Economics* (*N*=14), or in interdisciplinary ones such as *Journal of Economic Psychology* (*N*=8). The remaining papers where published in journals concerning different topics, such as biology (*N*=6), mathematics and game theory (*N*=4), sustainability (*N*=3), or commons (*N*=2).

@@

Finally, most of the studies were conducted in Europe (*N*=40) (especially in Germany (*N*=12) and in the Netherlands (*N*=8)), followed by North America (*N*=20, of which *N*=19 in the USA) and Asia (*N*=7, of which *N*=6 in China). Moreover, six studies were set in more than one country, and fourteen were meta-analysis or literature reviews. The two remaining studies took place in Australia and Colombia, respectively.

@@

Our analysis identified five individuals factors that impact cooperation: gender, social status, group identification, values and personality traits, and trust. In the following sections, we describe their principal effects on cooperation, and provide an overview of the moderators of each effect. More details about moderators are included in Appendix A.

@@

The effect of gender on cooperation is controversial, and studies that investigated it provided contradictory results. Sometimes men cooperated more [@Dorrough2019], while sometimes women did [@Peshkovskaya2017]. However, two meta-analyses reported a non significant effect size [@Balliet2011; @Spadaro2023]. This indicates that gender differences in cooperation are very flexible and depend on the context [@Spadaro2023]. More specifically, @Sell2009 explained these contradictory results "by structural differences [of the experimental settings] and identities triggered by those differences" (p. 322).

@@

Lastly, gender differences in cooperation could be partly explained by social status, where women represent the low-status group [@Sell2009]. For this reason, it is important to gain insights into how social status impacts cooperation.

@@

The results of studies investigating the impact of social status on cooperation do not allow us to draw a clear conclusion about this effect. For instance, some findings suggest that low-status participants cooperate more than high-status ones by contributing a higher proportion of their endowments [@Kingsley2016]. However, @VanLange2013 identified some studies where wealthier people cooperated more.

Social status can also create inter-class conflicts, because people tend to cooperate more with the members of the same social class than with members of other classes [@Aksoy2019]. This is especially true for low-status groups, where solidarity can arise either in the form of discrimination toward high-status members [@Camera2020], or in the form of fewer punishments for other low-status [@Chen2022].

Regarding the effect of overall group composition, studies disagree. On one hand, @Aksoy2019 found that groups with an inequality of status between members displayed lower levels of cooperation than homogeneous groups. On the other hand, economic and socio-cultural heterogeneity are only negatively correlated with in-group trust, while there is no evidence of their effect on the integrity of a common-pool resource [@VanKlingeren2021].

Taken together, these results suggest that the impact of social status on cooperation is complex. Moreover, similarly to what @Sell2009 reported about the effects of gender and social status, @Aksoy2019 demonstrated that the lower level of cooperation between members of different social classes is mainly due to actors' identification with their classes. This shows that social identity and social status are deeply connected, and that it is difficult to separate their effects on cooperation.

@@

In general, people cooperate more with other in-group than with out-group members [@Dorrough2015; @Aksoy2019], but the individual differences in the strength of identification can lead to inequalities in cooperation. In fact, individuals who strongly identify with their group can even compensate, by cooperating more, for other in-group members' defection [@VanVugt2009; @Arora2016].

Intragroup cooperation can be enhanced via a competition between different groups, because in such a context individuals' and groups' interests are aligned [@Puurtinen2009]. Moreover, a study showed that intergroup competition increased cooperation within the group, while individual, dispositional competitiveness had the opposite effect [@Nockur2020]. This demonstrated that the increase of cooperation via competition is due to group membership factors, and not to individuals' competitiveness.

Despite the beneficial effects of inducing intergroup competition, this could be harmful in real-life situations, where resources are shared between several groups. In fact, there is a greater risk of over-exploitation when an intergroup competition is present [@VanVugt2009]. To maintain the positive effect of group identification while eliminating the risks associated with competition, it is useful to prime a super-ordinate identity [@VanVugt2009]. This can be done by making salient the common characteristics of the different groups rather than their differences [@Kramer1984] or by inducing the feeling of a common fate [@Zhang2019].

@@

Trust can be defined as the "expectations of others' benevolent motives in situations that involve a conflict between self and collective interests" [@Balliet2013, p.1]. There are two categories of trust: first, dispositional trust is an individual general trait that can be assessed with a personality scale. Secondly, state trust is linked to the context and represent the expectations about a partner's behaviour in a specific situation [@Balliet2013]. The two are positively correlated [@Lubke2021], and both have a positive effect on cooperation [@VanLange2013; @Balliet2013].

Several studies have confirmed this relationship. For instance, high levels of trust are associated with more contributions to a public good [@Kocher2015], individuals' cooperation can be predicted by their beliefs about others' contributions [@Smith2013], people who are pessimistic about others' contribution cooperate significantly less than optimistic people [@Oyediran2018], and individuals who cooperate more display higher levels of trust in following interactions [@Chaudhuri2002]. The enhancing effect of trust on cooperation has also been demonstrated in real-life studies [see for example @Sturm2019; @Franzen2019]. Finally, trust matters more in situations involving a large, rather than small, conflict between personal and collective interests [@Balliet2013], and when people are uncertain about others' intentions [@VanLange2013].

People can also display different levels of trust in the "reliability, effectiveness and legitimacy of public institutions" [@Rompf2017, p. 2]. Institutional trust is associated with the probability of recycling, because knowing that sanctions can be efficiently applied motivates people to comply with norms.

@@

On the organizational level, we identified four factors from the reviewed papers: incentives, communication, social norms and anonymity. As for the individual factors, in the following sections we detail the main effects of each factor, and we provide an overview of possible moderators. More details can be found in Appendix A.

@@

Cooperation can be modified by introducing incentive systems, that can be either punishments or rewards. Both have a positive effect on cooperation [@Balliet2011a]. However, some studies found that punishments are more efficient than rewards [e.g. @Milinski], while others came to the opposite conclusion [e.g. @Chen2022].

Several factors moderate the main effect of incentives on cooperation. A meta-analysis found that both rewards and punishments are more effective if their implementation is costly rather than free, and during iterated, rather than single, interactions [@Balliet2011a]. Moreover, punishments work better in groups where all members have the same income [@Kingsley2016], because in heterogeneous groups a form of solidarity between low-status members arises and encourages free-riding [@Chen2022]. Punishments are also more effective if participants are friends [@Balliet2011a] and when only cooperators have the power to punish defectors [@Xiao2016]. Finally, the link between behaviour and outcome must be clear [@Xiao2016], and the punishment must have a significant consequence [@Egas2008].

In general, there are more studies on punishments than on rewards [@VanDijk2015]. When a punishment system is present, participants' perceive the situation as more fair [@Nockur2021]. However, punishments must be applied with carefulness. Ideally, groups should not punish in the early stages of the interaction, and have a norm of when and how to punish [@Rockenbach2019]. Moreover, a punishment system could entail a second-order social dilemma, since cooperators are better-off by not investing resources into punishing, if all others do [@Milinski2012]. Also, if there are no rules about who can punish, non-cooperators can punish cooperators as a form of retaliation [@Xiao2016; @VanDijk2015]. A democratic punishment system, where a fine is given only if a majority agree, could solve this problem [@Pfattheicher2018].

Despite enhancing cooperation, incentives have negative sides: they reduce participants' intrinsic motivation [@VanDijk2015; @Schlosser2018] by framing the situation as an economic, rather than ethical, dilemma [@VanVugt2009]. They also undermine participants' trust in other members' internal motivation to cooperate. Therefore, if they are removed, cooperation drops [@Irwin2014; @Mulder2006].

Finally, incentives are often studied in economic formats, but they can also be social. In fact, expecting honour or shame (after cooperating or defecting, respectively), or expressing disapproval toward defectors can enhance cooperation [@Jacquet2011; @Nelissen2013]. An ostracism system where participants can vote to exclude other members is also very effective [@Maier-Rigaud2010]. In conclusion, social sanctions can be even more efficient than economic ones, because cooperation do not drop after their removal [@Nelissen2013].

@@

Communication between members has a general positive effect on cooperation [e.g. @Balliet2010], because it allows to shape a cooperative social norms [@Janssen2014; @Bicchieri2002] and to reach a consensus about the best strategy for the group [@Hopthrow2005]. In fact, when communication is used to foster a common social identity, promote cooperative social norms and ensure that every member publicly commits, cooperation is higher [@Adams2022].

Concerning the moderation effects, oral communication is more powerful than written messages, and communication has a stronger impact in large, rather than small, groups [@Balliet2010]. Moreover, if communicating requires a personal cost, its effect is weaker than if communication is free [@Janssen2014]. The content of the message is also essential: to be effective, communication must allow participants to become aware of the problem faced by the group, identify different strategies and agree on the one to use. In addition, casual chit-chat also plays a very important role in increasing cooperation [@Koessler2021]. Finally, communication before or during the cooperative interaction have both a positive impact on cooperation [@Balliet2010], while knowing that it will be possible to communicate after the interaction does not enhance cooperation [@Kumakawa2013], and can even have a negative effect [@Torsvik2011].

A particular form of communication is gossip, i.e. the spread of reputational information about another person [@Feinberg2014]. Gossip increases cooperation by providing information about the past behaviour of potential partners, which helps to select the most cooperative members in the following interaction. Moreover, knowing that gossip is possible motivates people to cooperate more to avoid being the target of such gossip [@Feinberg2014]. Gossip can also have the pro-social function of protecting others from potential exploitation [@Feinberg2012]. Finally, providing correct information while gossiping helps to build a positive reputation [@Giardini2021] and reduces negative affects on the gossip actor [@Feinberg2012].

@@

Social norms can provide a guide on how to behave, especially when there is uncertainty about others' behaviour like in cooperative situations [@VonBorgstede2018]. Both injunctive (what others approve) and descriptive (how others actually behave) norms impact cooperative behaviours [@Thogersen2008]. Moreover, social norms help to maintain cooperation, because individuals who do not respect them are susceptible to social sanctions [@Rege2004].
